Clustering Beat-Chroma Patterns\\ in a Large Music Database

### 2010-03-24 Paper/Paper.tex
figure 1 changed, no grid
dpwe's changes up to section 2.2

\begin{document}
%
\maketitle
%
\begin{abstract}
A musical style or genre implies a set of common conventions
and patterns combined and deployed in different ways to make
individual musical pieces; for instance, most would agree that
contemporary pop music is assembled from a relatively small
palette of harmonic and melodic patterns.  The purpose of this
paper is to use a database of tens of thousands of songs
in combination with a compact representation of melodic-harmonic
content, the beat-synchronous chromagram, and data-mining
tools (clustering) to attempt to explicitly catalog this palette --
at least within the limitations of the beat-chroma representation.
We use online $k$-means clustering to summarize 3.7 million
4-beat bars in a codebook of a few hundred prototypes.
By measuring how accurately such a quantized codebook
can reconstruct the original data, we can quantify the degree
of diversity (distortion as a function of codebook size) and
temporal structure (\ie the advantage gained
by joint quantizing multiple frames) in this music.  The most
popular codewords themselves reveal the common chords
used in the music.  Finally, the quantized representation of
music can be used for music retrieval tasks such as artist
and genre classification, and identifying songs that are
similar in terms of their melodic-harmonic content.
\end{abstract}
%
\section{Introduction}\label{sec:introduction}
The availability of very large collections of music audio present
many interesting research opportunities.  Given millions of examples
from a single, broad class (\eg contemporary commercial pop music),
can we infer anything about the underlying structure and common
features of this class?  This paper describes our work in this direction.

What are the common features of pop music?  There are conventions
of subject matter, instrumentation, form, rhythm, harmony, and melody, among
others.  Our interest here is in the tonal content of the music -- \ie the
harmony and melody.  As a computationally convenient proxy for a richer
description of the tonal content of audio, we use the popular Chroma
representation, which collapses an acoustic spectrum into a 12-dimensional
description, with one bin for each semitone of the western musical octave.
In addition, we simplify the time axis of our representation to take advantage
of the strong beat present in most pop music, and record just one
chroma vector per beat.  This beat-synchronous chromagram
representation represents a typical music track
in a few thousand values, yet when resynthesized back into audio via
modulation of octave-invariant ``Shepard tones'', the melody and
chord sequences of the original music typically remain recognizable
\cite{Ellis2007a}.  To the extent, then, that beat-chroma
representations preserve tonal structure, they
are an interesting domain in which to search for patterns -- rich enough
to generate musically-relevant results, but simplified enough to
abstract away some other aspects of the original audio, such as
instrumentation and other stylistic details.

Specifically, this paper identifies common patterns in beat-synchronous
chromagrams by learning codebooks from a large set of examples.
The individual codewords consist of short beat-chroma patches, typically
between 1 and 8 beats, which are able to represent the entire dataset
of millions of patterns with the minimum error given a small codebook of
a few hundred entries.  Our goal is to identify meaningful information
about the musical structure represented in the entire database by
examining individual entries in this codebook.  Since the common
patterns represent a higher-level description of the musical content,
we also expect them to be useful in other applications, such as
music classification and retrieving tonally-related items.

One major issue in working with small patches concerns temporal
alignment. Learning the same motifs with different offset
(of a few seconds, or audio frames) is redundant. We take advantage
of rhythmic knowledge about the song to align patches with beats.
We can further group beats into bars to produces patches
for meaningful music segments.
The use of the Echo Nest API \cite{EchoNest} for that purpose is 
explained in Subsection \ref{ssec:beatpatch}. Patches built around
those temporal landmarks should capture music motifs more easily.

Prior work using small patches of chroma features includes
the ``shingles'' of \cite{Casey2007}, which they used to
identify ``remixes'' i.e. music based on some of the same
underlying instrument tracks, as well as for matching
performances of Mazurkas \cite{Casey2008}.  That work,
however, was not concerned with extracting the deeper
common patterns underlying different pieces (and did not
use either beat- or bar-synchronous features).  Earlier work in
beat-synchronous analysis includes \cite{Bartsch2001},
who looked for repeated patterns within single songs to
identify the chorus, and \cite{Ellis2007a}, who
cross-correlated beat-chroma matrices to match cover
versions of pop music tracks.  None of these works
examined or interpreted the content of the chroma
matrices in any detail.  In contrast, here we hope to
develop a codebook whose entries are of interest
in their own right.

@@

\subsection{Features}
The feature analysis used throughout this work is based on Echo Nest
analyze API \cite{EchoNest}.  
%
For any song uploaded to their platform this analysis returns a chroma
vector (length $12$) for every music event (called ``segment''), and a
segmentation of the song into beats and bars. Beats span multiple
segments, and bars span multiple beats.
%
Averaging the per-segment chroma over beat times results in a
beat-synchronous chroma feature representation similar to that used in
\cite{Ellis2007a}.

@@

Note that we do not claim that any of these informations (segments, beats, bars)
are perfectly accurate (\eg in \cite{Barrington2009a} authors obtain a better
song segmentation than the Echo Nest). 
Practice showed us that they are reasonable, 
and the size of the data set should make up for the imperfections or noise.
We also believe that patches sizes based on a number of beats or bars are more
meaningful than an arbitrary time length. More on this in the experiments
(Subsection \ref{ssec:segment}).

@@

We use the bar segmentation obtained from the Echo Nest analysis to
break a song into a collection of beat-chroma ``patches'', typically
one or two bars in length.
%
Because the bar length varies, depending on the meter of a particular
song, we resample each patch to a fixed length of typically 4 beats
per bar.  Since the overwhelming majority (82\%) of our training data
consisted of bars that were 4 beats long, this resampling was not
often required.  Most of the remaining bars (10\%) were 3 beats in
length.
% although the majority of the data contains bars that are 4 beats long
The resulting patches consist of $12 \times 4$ or $12 \times 8$ matrices.

Finally, we normalize the patches with respect to the key by rotating
the pattern matrix so that the first row contains the most
energy. This can easily be seen in Figure \ref{fig:code}.
% We considered the same technique for rolling along the other axis and
% be invariant regarding the downbeat. However, it seems less robust to
% noise.
Each patch within a song is normalized independently, so exact
reconstruction of the original song requires knowledge of the
rotation index (key) for each patch.

The representation resulting from this process is invariant to both
the key and meter of the original song.  This enables the study of
broad harmonic patterns found throughout the data, without regard for
the specific musical context.
%
% In the context of clustering, this enables us to study harmonic
% pattern usage without maintaining clusters for the specific context a
% particular patterns occurs in,
In the context of clustering this avoids \eg obtaining separate
clusters for every major triad in duple or triple meter.
% Also, this implies that the pitch envelope is important rather than
% the actual pitches.
%FIXME: Something about why we're not interested in specific chords?
Specific chords do not add much information about a music collection.
A cover song can be performed in a different key, but the motifs it
contains remain the same.

>>>>

\begin{document}
%
\maketitle
%
\begin{abstract}
A musical style or genre implies a set of common conventions
and patterns combined and deployed in different ways to make
individual musical pieces; for instance, most would agree that
contemporary pop music is assembled from a relatively small
palette of harmonic and melodic patterns.  The purpose of this
paper is to use a database of tens of thousands of songs
in combination with a compact representation of melodic-harmonic
content (the beat-synchronous chromagram) and data-mining
tools (clustering) to attempt to explicitly catalog this palette --
at least within the limitations of the beat-chroma representation.
We use online $k$-means clustering to summarize 3.7 million
4-beat bars in a codebook of a few hundred prototypes.
By measuring how accurately such a quantized codebook
can reconstruct the original data, we can quantify the degree
of diversity (distortion as a function of codebook size) and
temporal structure (\ie the advantage gained
by joint quantizing multiple frames) in this music.  The most
popular codewords themselves reveal the common chords
used in the music.  Finally, the quantized representation of
music can be used for music retrieval tasks such as artist
and genre classification, and identifying songs that are
similar in terms of their melodic-harmonic content.
\end{abstract}
%
\section{Introduction}\label{sec:introduction}
The availability of very large collections of music audio present
many interesting research opportunities.  Given millions of examples
from a single, broad class (\eg contemporary commercial pop music),
can we infer anything about the underlying structure and common
features of this class?  This paper describes our work in this direction.

What are the common features of pop music?  There are conventions
of subject matter, instrumentation, form, rhythm, harmony, and melody, among
others.  Our interest here is in the tonal content of the music -- \ie the
harmony and melody.  As a computationally-convenient proxy for a richer
description of the tonal content of audio, we use the popular Chroma
representation, which collapses an acoustic spectrum into a 12-dimensional
description, with one bin for each semitone of the western musical octave.
In addition, we simplify the time axis of our representation to take advantage
of the strong beat present in most pop music, and record just one
chroma vector per beat.  This beat-synchronous chromagram
representation represents a typical music track
in a few thousand values, yet when resynthesized back into audio via
modulation of octave-invariant ``Shepard tones'', the melody and
chord sequences of the original music usually remain recognizable
\cite{Ellis2007a}.  To the extent, then, that beat-chroma
representations preserve tonal content, they
are an interesting domain in which to search for patterns -- rich enough
to generate musically-relevant results, but simplified enough to
abstract away aspects of the original audio such as
instrumentation and other stylistic details.

Specifically, this paper identifies common patterns in beat-synchronous
chromagrams by learning codebooks from a large set of examples.
The individual codewords consist of short beat-chroma patches of
between 1 and 8 beats, which are able to represent the entire dataset
of millions of patterns with minimum error given a small codebook of
a few hundred entries.  Our goal is to identify meaningful information
about the musical structure represented in the entire database by
examining individual entries in this codebook.  Since the common
patterns represent a higher-level description of the musical content
than the raw chroma,
we also expect them to be useful in other applications, such as
music classification and retrieving tonally-related items.

One major issue in working with small patches concerns temporal
alignment. Learning the same motifs with different offsets 
(in time or beats) is redundant. We take advantage
of rhythmic knowledge about the song to align patches with beats.
We can further group beats into bars to produces patches
for meaningful music segments.
Our use of the Echo Nest API \cite{EchoNest} for this purpose is 
explained in Section \ref{ssec:beatpatch}. Patches built around
these temporal landmarks should capture music motifs more easily.

Prior work using small patches of chroma features includes
the ``shingles'' of \cite{Casey2007}, which were used to
identify ``remixes'', i.e., music based on some of the same
underlying instrument tracks, and also for matching
performances of Mazurkas \cite{Casey2008}.  That work,
however, was not concerned with extracting the deeper
common patterns underlying different pieces (and did not
use either beat- or bar-synchronous features).  Earlier work in
beat-synchronous analysis includes \cite{Bartsch2001},
which looked for repeated patterns within single songs to
identify the chorus, and \cite{Ellis2007a}, which 
cross-correlated beat-chroma matrices to match cover
versions of pop music tracks.  None of these works
examined or interpreted the content of the chroma
matrices in any detail.  In contrast, here we hope to
develop a codebook whose entries are of interest
in their own right.

@@

\subsection{Features}
The feature analysis used throughout this work is based on Echo Nest
analyze API \cite{EchoNest}.  
%
For any song uploaded to their platform this analysis returns a chroma
vector (length $12$) for every music event (called ``segment''), and a
segmentation of the song into beats and bars. Beats may span or 
subdivide segments; bars span multiple beats.
%
Averaging the per-segment chroma over beat times results in a
beat-synchronous chroma feature representation similar to that used in
\cite{Ellis2007a}.  Echo Nest chroma vectors are normalized to have the 
largest value in each column equal to 1.

@@

Note that none of this information (segments, beats, bars)
can be assumed perfectly accurate.
%(\eg in \cite{Barrington2009a} authors obtain a better
%song segmentation than the Echo Nest). 
In practice, we have found them reasonable, 
and given the size of the data set, any rare imperfections or noise
can be diluted to irrelevance by the good examples.  
We also believe that patch sizes based on a number of beats or bars are more
meaningful than an arbitrary time length. This is discussed further in 
Section \ref{ssec:segment}.

@@

We use the bar segmentation obtained from the Echo Nest analysis to
break a song into a collection of beat-chroma ``patches'', typically
one or two bars in length.
%
Because the bar length is not guaranteed to be 4 beats, 
depending on the meter of a particular
song, we resample each patch to a fixed length of typically 4 beats
per bar.  However, the majority (82\%) of our training data
consisted of bars that were 4 beats long, so this resampling 
usually had no effect.  Most of the remaining bars (10\%) were 3 beats in
length.
% although the majority of the data contains bars that are 4 beats long
The resulting patches consist of $12 \times 4$ or $12 \times 8$ matrices.

Finally, we normalize the patches with respect to transposition by rotating
the pattern matrix so that the first row contains the most
energy. This can be seen in the example codeword of Figure \ref{fig:code}.
% We considered the same technique for rolling along the other axis and
% be invariant regarding the downbeat. However, it seems less robust to
% noise.
Each patch within a song is normalized independently, so
reconstruction of the original song requires knowledge of the
rotation index (key) for each patch.

The representation resulting from this process is invariant to both
the key and meter of the original song.  This enables the study of
broad harmonic patterns found throughout the data, without regard for
the specific musical context.
%
% In the context of clustering, this enables us to study harmonic
% pattern usage without maintaining clusters for the specific context a
% particular patterns occurs in,
In the context of clustering this avoids problems such as obtaining separate
clusters for every major triad in duple or triple meter.
% Also, this implies that the pitch envelope is important rather than
% the actual pitches.
%FIXME: Something about why we're not interested in specific chords?
%
%Specific chords do not add much information about a music collection.
%A cover song can be performed in a different key, but the motifs it
%contains remain the same.
%%%% dpwe: I don't agree with this, I'd rather say nothing
